The relation betweem feeling and expressing emotions. A comparison between sujective feeling and facial expression recognition of spontaneous emotions.

### 2019-03-15 ACII_2019_paper/ACII_2019_paper.tex
acii paper

@@

% Pandoc header

@@

With the developement of commercial automatic facial expression
recognition tools, industries and governments are gradually implementing
this technology in order to track humans' emotions in various scenarios
(e.g., marketing, heathcare, automotive to name a few). Beside the
ethical question of measuring human emotions, these tools are
challenging psychological theories about infering emotions from facial
expressions. When automatic facial expression recognition tools are used
to measure human emotions, they rest on the premise that facial
expressions provide a direct access to expressers' emotions. However the
link between emotion felt and facial expressions is far from being
established and still remain a hot topic in psychology research.

@@

After their emotion eliciation task the participants had to rate their
emotional state during the task on a likert scale from 0 (``not at
all'') to 5 (``strongly'') the six ``basic'' emotions (i.e.,
\emph{Anger}, \emph{Disgust}, \emph{Fear}, \emph{Happiness},
\emph{Surprise} and \emph{Sadness}) as well as six ``non-basic''
emotions (i.e., \emph{Pride}, \emph{Curiosity}, \emph{Boredom},
\emph{Shame}, \emph{Humiliation}, and \emph{Disappointement}). The
emotion scales rated as the highest is considered as describing
participants subjective feeling.

@@

The annotation of facial expressions was performed on-site using
\emph{Oudjat}, a software for designing video annotation experiements
(Dupré et al. 2015). For each video, the annotation procedure hat two
steps. First, the participants had to identify the emotional sequences
by pressing the space bar of their keyboard to indicate the begining and
the end of the emotional sequences while watching the video. Second, the
participants watched each emotional sequence previously identified and
had to label the sequence using one of the 12 emotions proposed
including six ``basic'' emotions (i.e., \emph{Anger}, \emph{Disgust},
\emph{Fear}, \emph{Happiness}, \emph{Surprise} and \emph{Sadness}) and
six ``non-basic'' emotions (i.e., \emph{Pride}, \emph{Curiosity},
\emph{Boredom}, \emph{Shame}, \emph{Humiliation}, and
\emph{Disappointement}). They also had the possibility to indicate that
the sequence was not expressing one of the proposed emotion.

This annotation procedure results in a uni-dimesional time-serie for
each video per human observer identifying for each second of the video
which emotion was recognized. Then time-series corresponding to the same
video were aggreated to calculate the proprotion of human observers for
each second of the video per emotional label. The sum of each label
proportion per second was used as a score to determine which labels
corresponds to the overall video (i.e., the highest score).

@@

The 232 annotated video were processed with Affdex (SDK v3.4.1). Affdex
is an automatic facial expression recognition classifier developed and
distributed by Affectiva is a spinoff company resulting from the
research activities of MIT media lab created in 2009 (McDuff et al.
2016). Affdex's algorithm uses Histogram of Oriented Gradient (HOG)
features and Support Vector Machine (SVM) classifiers in order to
recognize facial expressions. For each video frame, Affdex identify the
probability of the face as expressing one of the six ``basic'' emotions
(i.e., \emph{Anger}, \emph{Disgust}, \emph{Fear}, \emph{Happiness},
\emph{Surprise} and \emph{Sadness}) as well as additional psychological
states such as \emph{Valence}, \emph{Engagement} or \emph{Contempt}, and
facial features such as \emph{Cheek Raise}, \emph{Eye Widen} or
\emph{Jaw Drop}.

@@

Whereas the self-reports, the human annotations and the automatic
recognition include data on ``non-basic'' emotions and features, the
analysis is performed using only the six ``basic'' emotions in order to
compare them.

@@

@@

@@

The conclusion goes here.

@@

@@

>>>>

\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}

@@

% Pandoc header
\usepackage{booktabs}
\usepackage{float}
\usepackage{tabu}

@@

With the developement of commercial automatic facial expression
recognition tools (see Dupré et al. 2018 for a non-exhaustive list of
available tools), industries and governments are gradually implementing
this technology in order to track humans' emotions in various scenarios
(e.g., marketing, heathcare, automotive to name a few). Beside the
ethical question of measuring human emotions, these tools are
challenging psychological theories about infering emotions from facial
expressions. When automatic facial expression recognition tools are used
to measure human emotions, they rest on the premise that facial
expressions provide a direct access to expressers' emotions. However the
link between emotion felt and facial expressions is far from being
established and still remain a hot topic in psychology research.

@@

After their emotion eliciation task the participants had to rate their
emotional state during the task on a likert scale from 0 (``not at
all'') to 5 (``strongly'') the six ``basic'' emotions (i.e.,
\emph{anger}, \emph{disgust}, \emph{fear}, \emph{happiness},
\emph{surprise} and \emph{sadness}) as well as six ``non-basic''
emotions (i.e., \emph{pride}, \emph{curiosity}, \emph{boredom},
\emph{shame}, \emph{humiliation}, and \emph{disappointement}).

@@

The annotation of facial expressions was performed on-site using
\emph{Oudjat}, a software for designing video annotation experiements
(Dupré et al. 2015). For each video, the annotation procedure hat two
steps. First, the participants had to identify the emotional sequences
by pressing the space bar of their keyboard to indicate the begining and
the end of the emotional sequences while watching the video. Second, the
participants watched each emotional sequence previously identified and
had to label the sequence using one of the 12 emotions proposed
including six ``basic'' emotions (i.e., \emph{anger}, \emph{disgust},
\emph{fear}, \emph{happiness}, \emph{surprise} and \emph{sadness}) and
six ``non-basic'' emotions (i.e., \emph{pride}, \emph{curiosity},
\emph{boredom}, \emph{shame}, \emph{humiliation}, and
\emph{disappointement}). They also had the possibility to indicate that
the sequence was not expressing one of the proposed emotion.

This annotation procedure results in a uni-dimesional time-serie for
each video per human observer identifying for each second of the video
which emotion was recognized. Then time-series corresponding to the same
video were aggreated to calculate the proprotion of human observers for
each second of the video per emotional label. The sum of each label
proportion per second was used as a score to determine which labels
corresponds to the overall video (i.e., the highest score). In case of
more than one label having the maximum value, the emotion is discribed
as undetermined.

@@

The 232 annotated video were processed with Affdex (SDK v3.4.1). Affdex
is an automatic facial expression recognition classifier developed and
distributed by Affectiva is a spinoff company resulting from the
research activities of MIT media lab created in 2009 (McDuff et al.
2016). Affdex's algorithm uses Histogram of Oriented Gradient (HOG)
features and Support Vector Machine (SVM) classifiers in order to
recognize facial expressions. For each video frame, Affdex identify the
probability of the face as expressing one of the six ``basic'' emotions
(i.e., \emph{anger}, \emph{disgust}, \emph{fear}, \emph{happiness},
\emph{surprise} and \emph{sadness}) as well as additional psychological
states such as \emph{valence}, \emph{engagement} or \emph{contempt}, and
facial features such as \emph{cheek raise}, \emph{eye widen} or
\emph{jaw drop}.

@@

Whereas the self-reports, the human annotations and the automatic
recognition include data on ``non-basic'' emotions and features, the
analysis is performed using only the six ``basic'' emotions in order to
compare them. The maximum score for self-reports, human annotations and
automatic recognition is used to label the video. In case of more than
one label obtaining the maximum value, the video is labeled as
undetermined.

@@

Emotions self-reported as being characteristic of the elicitation are
compared with the emotion recognized by the human observers in a
confusion matrix (Figure \ref{fig:confusionMatrix_sr_hr}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_sr_hr-1.pdf}
\caption{Confusion matrix of between the emotion self-reported as being
characteristic of the elicitation with the emotion recognized by the
human observers.}
\end{figure}

The result of the confusion matrix show a low agreement between emotion
felt during the elecitation and emotion recognized by the human
annotators (Accuracy = 0.27, 95\%CI{[}0.21,0.33{]}; Kappa = 0.11) except
for \emph{happiness} (15.2\%), \emph{surprise} (8.26\%) and
\emph{disgust} (1.74\%). Sensitivity, specificity, precision and F1
score for each emotion can be found Table
\ref{tab:confusionTable_sr_hr}. Interestingly human annotators seem to
recognize as \emph{surprise} videos in which \emph{happiness} was the
highest self-reported emotion (10.9\%), and in a lower instance
\emph{happiness} videos in which \emph{surprise} was the highest
self-reported emotion (4.78\%).

\begin{table}[!h]

\caption{\label{tab:confusionTable_sr_hr}Agreement accuracy metrics for each emotion. }
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
Emotion & Sensitivity & Specificity & Precision & F1\\
\midrule
anger & 0.43 & 0.92 & 0.14 & 0.21\\
disgust & 1.00 & 0.93 & 0.20 & 0.33\\
fear & 0.00 & 0.96 & 0.00 & \textit{na.}\\
happiness & 0.49 & 0.73 & 0.45 & 0.47\\
sadness & 0.08 & 0.91 & 0.05 & 0.06\\
surprise & 0.44 & 0.67 & 0.23 & 0.31\\
undetermined & 0.00 & 0.99 & 0.00 & \textit{na.}\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} Note. \textit{na.} values are produced when not enough}\\
\multicolumn{5}{l}{data are available to compute accuracy indicators.}\\
\end{tabu}
\end{table}

However, the self-report show a very high proportion of undetermined
emotional states which reveals not only the possiblility of the emotion
elicitation tasks to trigger more than one emotion but also the
potential limit of using 6-points likert scales for which the
participants can easily score to the maximum for more than one emotion.

@@

As in the previous analysis, emotions self-reported as being
characteristic of the elicitation are compared with the emotion
recognized by the automatic classifier in a confusion matrix (Figure
\ref{fig:confusionMatrix_sr_ar}).

\begin{figure}
\centering
\includegraphics{ACII_2019_paper_files/figure-latex/confusionMatrix_sr_ar-1.pdf}
\caption{Confusion matrix of between the emotion self-reported as being
characteristic of the elicitation with the emotion recognized by the
automatic classifier.}
\end{figure}

Results obtained for the comparison between emotions self-reported and
recognised by the automatic classifier are similar to the ones with
human observers (Table \ref{tab:confusionTable_sr_ar}). Overall there is
a low agreement between emotion self-reported and emotion recognized by
the automatic classifier (Accuracy = 0.19, 95\%CI{[}0.14,0.24{]}; Kappa
= 0.05) except for \emph{happiness} (13.9\%) and \emph{surprise}
(3.48\%). Surprisingly the automatic classifier incorrectly recognized
as \emph{disgust} an important proportion of videos in which
\emph{happiness} was the highest self-reported emotion (7.83\%). In
parallel, the automatic classifier recognised as \emph{happiness} and
\emph{disgust} videos in which \emph{surprise} was the highest
self-reported emotion (respectively 5.22\% and 6.09\%).

\begin{table}[!h]

\caption{\label{tab:confusionTable_sr_ar}Agreement accuracy metrics for each emotion.}
\centering
\fontsize{8}{10}\selectfont
\begin{tabu} to \linewidth {>{\raggedright}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
Emotion & Sensitivity & Specificity & Precision & F1\\
\midrule
anger & 0.00 & 0.89 & 0 & \textit{na.}\\
disgust & 0.25 & 0.70 & 0.01 & 0.03\\
fear & 0.00 & 0.93 & 0 & \textit{na.}\\
happiness & 0.44 & 0.71 & 0.41 & 0.43\\
sadness & 0.15 & 0.94 & 0.14 & 0.15\\
surprise & 0.19 & 0.89 & 0.28 & 0.22\\
undetermined & 0.00 & 1.00 & \textit{na.} & \textit{na.}\\
\bottomrule
\multicolumn{5}{l}{\textsuperscript{} Note. \textit{na.} values are produced when not enough data are}\\
\multicolumn{5}{l}{available to compute accuracy indicators.}\\
\end{tabu}
\end{table}

A comparable explaination can be provided as the level of undetermined
emotion are very high for the self reports.

@@

Despit being one on the most investigated question in affective science,
the link between emotion felt and facial expression is a hot topic and
no clear evidence have been found to definively answer it. However, with
the growing interest of industries and govenment to monitor individual's
psychological states, evidences are showing that facial expressions are
in reality not expressing emotions (McKeown 2013). This research aimed
to provide some empirical data to the question. The subjective feeling
of participants was compared with human recognition on one side and
automatic recognition on the other side. The results reveals a low
accuracy for both humans and automatic classifier to accurately identify
the inner emotional states of these individuals based on their facial
expressions.

Some limitations to this process should be stated over the use of
self-reports to evalute individual's subjective feelings. Accessing to
the inner subjective feeling can be biaised if not impossible. Moreover
the laboratory setting can trigger ambiguous and ``non-basic'' emotions
which were not analysed in this research. The procedure use for human
annotation can also be incriminated. Instead of asking the human
annotators to provide an unique label, a more sutble approach was
choosen to mimic results provided by the automatic classifier. In this
regard, the results of the human annotation could have been more
ambiguous because it is not the natural way that people are infering
human emotions. Finaly, the automatic classifer algorithm can also be
problematic. Based on training datasets which are most of the time using
prototypical facial expression of the ``basic'' emotions, the algorithm
to classify facial expressions can be held in check by the spontanous
facial expressions analysed.

Considering the above, the results provides an additional evidence that
individuals' subjective feeling can not be infered from facial
expressions and in our case invalidate the hypothesis of hardwired
emotions. This result suggests that automatic facial expression
recognition tools should be focused on evaluating facial morphology
features such as action units rather than infering potentional emotional
or affective states.

@@

\leavevmode\hypertarget{ref-dupre2018accuracy}{}%
Dupré, Damien, Nicole Andelic, Gawain Morrison, and Gary McKeown. 2018.
``Accuracy of Three Commercial Automatic Emotion Recognition Systems
Across Different Individuals and Their Facial Expressions.'' In
\emph{2018 Ieee International Conference on Pervasive Computing and
Communications Workshops (Percom Workshops)}, 627--32. IEEE.

@@

\leavevmode\hypertarget{ref-mckeown2013analogical}{}%
McKeown, Gary J. 2013. ``The Analogical Peacock Hypothesis: The Sexual
Selection of Mind-Reading and Relational Cognition in Human
Communication.'' \emph{Review of General Psychology} 17 (3): 267--87.
